A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 MS.tex
Some references added and read through to `Number of gonotrophic cycles estimates from dissection' subsection

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.4
days (median estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 7.6 days (median estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.2 days (median estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 6.2 days (median estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 4.3 days (median estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 11.6 days
(median estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (median estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

There are two main strategies to estimate mosquito mortality rates and
longevity. The first is through mark-release-recapture (MRR)
experiments, a technique that is widely applied to estimate these
parameters in many types of animal. As applied to mosquitoes, insects
are caught in the field or reared in the laboratory and then marked,
typically with fluorescent dust. The mosquitoes are then released into
the field and then efforts are made to recapture them, for example using
human baits or light traps, usually over an extended period of time.
Mortality rates can be statistically estimated from the numbers of
recaptures given certain assumptions. The main challenges with MRR is
ensuring the marking technique does not affect recapture probability,
and distinguishing mortality from mosquitoes dispersing out of range of
being recaptured. Also, releasing insects that can transmit disease
(especially if this increases ambient population levels) raises
important ethical issues.

The second technique is specific to mosquitoes and makes use of their
gonotrophic cycle. The mosquito ovary is made up of ovarioles, each of
which typically produces one egg every gonotrophic cycle. After the egg
passes into the oviduct the distended ovariole does not completely
recover its previous form but a discrete dilation remains which can be
detected by dissecting the female reproductive organs. Data on the
fraction of females that have oviposited provides some information about
mortality rates. However, a skilled dissector can distinguish the number
of dilations from multiple gonotrophic cycles so providing much richer
data on longevity. The challenges of this method include the amount of
time and expertise it takes to collect the data, establishing the
relationships between physiological and chronological time (though the
distribution of the number of gonotrophic cycles wild-caught mosquitoes
have gone through is of direct epidemiological relevance) and the fact
that it only applies to females.

@@

In recent years many important vectors of disease have been shown to be
complexes of very closely related species, biotypes or forms that cannot
be distinguished morphologically (for example the morphospecies
\emph{Anopheles gambiae sensu lato} is now separated into the widespread
\emph{gambiae, coluzzii, arabiensis} and a number of more local
species). As the majority of studies analysed here took place before
molecular techniques allowed these taxa to be separated we work here
chiefly with morphospecies.

@@

Data from MRR experiments in the Guerra et al. (2014) database were
examined and those with fewer than six recaptures and species with only
a single MRR study were excluded for the hierarchical analysis. Of the
232 data sets, 177 involved only females, 35 males, and 18 both sex
releases. For 102 data sets the age of the released mosquitoes was known
(the average age of released mosquitoes was 4.0 days) while in the other
cases it was unknown or unrecorded; in these cases we assumed the
mosquitoes were newly emerged at the time of release and return to this
assumption later.

We analysed all MRR experiments within the same statistical framework
(for full details see the Supplementary Online Material). In the
simplest case \emph{N\textsubscript{R}} mosquitoes are released at day
zero and the probability that they remain in the recapture area until
day \emph{t} is \emph{S}(\emph{t}) when they are recaptured with
probability $\psi$. We model the number of mosquitoes recaptured on
day \emph{t} using a negative binomial sampling model with mean
\(\left( N_{R} - Y\left( t - 1 \right) \right)S\left( t \right)\psi\),
where \(Y\left( t - 1 \right)\) is cumulative captures before day
\emph{t}, and shape parameter $\kappa$. The negative binomial has been
used previously in MRR analyses (ref) because of its ability to
represent temporal over-dispersion in recaptures most likely caused by
variable weather. A slight modification was required for studies with
multiple releases (see SOM).

The simplest model for $S(t)$ assumes there is a constant
probability ($\lambda$) that a mosquito dies or leaves the recapture area
so that the numbers remaining after time \emph{t} are given by the
exponential distribution, \(exp\left\lbrack - \lambda t \right\rbrack\).
We utilised this form extensively but in testing for senescence used
five other models where $\lambda(t)$ varies with time so that

\begin{longtable}[]{@{}ll@{}}
\(S\left( t \right) = e^{- \int_{0}^{t}{\lambda\left( t \right) \mathrm{d}\tau}}.\)
\end{longtable}

@@

Parameters were estimated using Bayesian techniques with relative
uninformative priors for $\kappa$ and the parameters of
$\lambda(t)$, but assuming a prior for $\psi$ indicating a low
recapture probability (bounded in part by knowledge of the maximum daily
recapture rates). We used a Bayesian hierarchical model to estimate
distributions of lifespan at the species and the genus levels, and
across the complete data set. This procedure assumes that there is a
distribution of lifespan parameters for each species from which those
governing individual MRR time series are sampled, and similarly a
distribution at the genus level from which those for individual species
are derived (rather akin to random effects in classical statistics).
Within this framework we can also allow the parameters for individual
time series to be influenced by co-variates such as differences in
experimental methodology. As in the estimation of the parameters of the
individual experiments, relative uninformative priors were set for the
parameters of the hierarchical models except for $\psi$ where again a
distribution representing low recapture probabilities was assumed.
Posterior distributions were derived using Markov Chain Monte Carlo
(MCMC) methods with convergence assessed using the \(\hat{R}\) statistic
(Gelmin \& Rubin 1992). The predictive power of the model was assessed
using \emph{K}-fold cross validation which tests the ability of the
model fitted to part of the data to predict the rest using multiple
different partitions. Further details of the prior specification,
fitting and validation through posterior predictive checks are given in
the SOM.

@@

The Guerra et al. database included the latitude and longitude of each
study, along with the date when the study began. We used this information to find
estimates of the air temperature for each study using the European Centre for Medium
Range Weather Forecasts' ERA Interim Daily historical database. For each
study we calculated the mean monthly temperature across a spatial area
of (latitude $\pm$ 1 degree, longitude $\pm$ 1 degree), for the month at which
each study was carried out. The records for this database begin in 1979,
which pre-dates the study date for 65 of our 232 MRR time-series. For
these time-series, we chose to estimate the air temperature by an
average of the corresponding monthly temperatures over the years
1979-89.

@@

Most dissection studies recorded the distribution of the number of
gonotrophic cycles in mosquito samples collected over a specific period
of time. Overall, we found 568 time series in 72 published articles.
Because seldom were sufficient ancillary data available data to analyse
trends over time, we aggregated the data from each time series. We
further omitted time series with fewer than 100 mosquitoes and for
species with only one data set leaving 131 studies of mosquitoes in the
genera \emph{Anopheles, Aedes}, \emph{Culex} and \emph{Mansonia}.

@@

We modelled the number of mosquitoes found by dissection to be of age
\emph{a} using the negative binomial distribution with mean
$\Psi S(a)$ and shape parameter $\kappa$, where
$\Psi$ is the product of the recruitment rate of adult mosquitoes,
which we assume is constant over time, and the probability of being
captured for dissection, and \emph{S}(\emph{a}) is the probability of
surviving until age \emph{a}. We used the number of females that have
yet to lay eggs (nulliparous) to estimate the recruitment rate as
described further in the SOM. Initial examination revealed that in some
data sets the number of nulliparous females was anomalously low,
something that has been noticed before (Gillies \& Wilkes 1965). As some
studies have suggested that the first gonotrophic cycle tends to be
longer than the subsequent ones, this is probably due to differences in
capture probability. In data sets where the fraction of nulliparous
females was less than 90\% the uniparous (completed on gonotrophic
cycle) we excluded the nulliparous observation.

@@

MRR estimates the length of time a mosquito remains alive and is still
in the area available for recapture. It thus provides a lower bound to
lifespan which we shall refer to as LBL. In 187 of the 230 MRR time
series the estimated LBL was less than 10 days (Fig. \ref{fig:mrr_lifespan_individualEstimates}). The smallest
estimate was \textless{}1 day for the Asian malaria vector
\emph{Anopheles subpictus s.l.} which is unfeasibly short and almost
certainly reflects dispersal out of the recapture zone or a violation of
the assumptions of our analyses. The longest estimate was 18.3 days for
the temperate species \emph{Aedes simpsoni s.l.} which is a vector of
yellow fever in Africa. There are multiple data sets for the most
important vector species such as \emph{Anopheles gambiae, Aedes aegypti}
and \emph{albopictus} and \emph{Culex tarsalis} all of which show
considerable variation. For example, there are 54 estimates of LBL for
\emph{A. aegypti} which range from 2.2 days to 38.3 days with a mean of
8.3 days and coefficient of variation of 0.7. There are significant
differences in LBL amongst species (ANOVA on median LBL controlling for
sex and pre-release feeding: \emph{F}\textsubscript{37,194} = 2.5,
\emph{p} \textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{38}\), \emph{p}\textless{}0.01).

The estimated median LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.5, 5.0 and 6.9 days respectively with an overall estimate of 4.6 days
(Fig. \ref{fig:mrr_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
(\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}) individually (Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}).

>>>>

Mosquitoes are responsible for more human deaths than any other
organism, yet we still know relatively little about their ecology.
Mosquito lifespan is a key determinant of transmission strength for the
diseases they vector, but the field experiments used to determine this
quantity -- mark-release-recapture (MRR) studies and wild-caught
dissection of female mosquitoes -- produce estimates with high
uncertainty. In this paper, we use Bayesian hierarchical models to
analyse a previously-published database of 232 MRR experiments and
compile then analyse a database of 131 dissection studies to produce the
first ever species- and genus-level estimates of mosquito lifespan. Due
to the assumptions required to analyse the field data, we term our
estimates lower bounds on lifespan (LBL). Notably, for the major African
malaria vector \emph{Anopheles gambiae s.l.}, we estimate LBLs of 4.5
days (mean estimate; 25\%-75\% CI: 3.8-5.1 days for unfed female
mosquitoes from the MRR analysis) and 9.5 days (mean estimate;
25\%-75\% CI: 5.2-11.0 days from the dissection analysis); and an LBL of
4.3 days (mean estimate; 25\%-75\%
CI: 3.6-4.8 days, only present in the MRR database) for the
predominantly East-African vector \emph{A. funestus s.l}. We estimate
LBLs of 7.0 days (mean estimate; 25\%-75\% CI: 4.5-8.5 days from the
MRR analysis) and 5.0 days (mean estimate; 25\%-75\% CI: 3.5-5.1 days
from the dissection analysis) for \emph{Aedes aegypti}; and 12.1 days
(median estimate; 25\%-75\% CI: 10.0-13.7 days from the MRR analysis)
for \emph{Ae.} \emph{albopictus} -- the predominant vectors of dengue
fever, chikungunya and Zika. Our estimates indicate that there is
significant variation in lifespan across species, with most variation
explained by differences between genera. In correspondence with
laboratory studies, we estimate that female mosquitoes outlive males by
0.9 days on average (mean estimate; 25\%-75\% CI: 0.3-1.6 days). We
fit models incorporating mosquito senescence to the data, which allows
us to assess evidence for age-dependent mortality in mosquitoes across
different species. We determine that 8 of 33 species included in the MRR
database indicated evidence for senescence, versus only 2 of 25 species
from the dissection database. Our analysis applies a common framework to
the analysis of databases of MRR and dissection-based experiments,
allowing us to produce robust estimates of lower bounds on lifespan. It
also enables us to critically appraise each field method, highlighting a need for alternative field methods for measuring this
important mosquito characteristic.

@@

There are two main strategies to estimate mosquito mortality rates and
longevity. The first is through mark-release-recapture (MRR)
experiments, a technique that is widely applied to estimate these
parameters in many types of animal. As applied to mosquitoes, insects
are caught in the field or reared in the laboratory and then marked,
typically with fluorescent dust. The mosquitoes are then released into
the field and then efforts are made to recapture them, for example using
human baits or light traps, usually over an extended period of time.
Mortality rates can be statistically estimated from the numbers of
recaptures given certain assumptions \citep{silver2007mosquito}. The main challenges with MRR is
ensuring the marking technique does not affect recapture probability,
and distinguishing mortality from mosquitoes dispersing out of range of
being recaptured. Also, releasing insects that can transmit disease
(especially if this increases ambient population levels) raises
important ethical issues.

The second technique is specific to mosquitoes and makes use of their
gonotrophic cycle \citep{polovodova1949determination,detinova1962age}. The mosquito ovary is made up of ovarioles, each of
which typically produces one egg every gonotrophic cycle. After the egg
passes into the oviduct the distended ovariole does not completely
recover its previous form but a discrete dilation remains which can be
detected by dissecting the female reproductive organs. Data on the
fraction of females that have oviposited provides some information about
mortality rates. However, a skilled dissector can distinguish the number
of dilations from multiple gonotrophic cycles so providing much richer
data on longevity. The challenges of this method include the amount of
time and expertise it takes to collect the data, establishing the
relationships between physiological and chronological time (though the
distribution of the number of gonotrophic cycles wild-caught mosquitoes
have gone through is of direct epidemiological relevance) and the fact
that it only applies to females.

@@

In recent years many important vectors of disease have been shown to be
complexes of closely related species, biotypes or forms that cannot
be distinguished morphologically (for example the morphospecies
\emph{Anopheles gambiae sensu lato} is now separated into the widespread
\emph{gambiae, coluzzii, arabiensis} and a number of more local
species). As the majority of studies analysed here took place before
molecular techniques allowed these taxa to be separated we work here
chiefly with morphospecies.

@@

Data from MRR experiments in the Guerra et al. (2014) database were
examined and those with fewer than six recaptures and species with only
a single MRR study were excluded for the hierarchical analysis. Of the
232 data sets, 179 involved only females, 35 males, and 18 both sex
releases. For 102 data sets the age of the released mosquitoes was known
(the average age of released mosquitoes was 4.0 days) while in the other
cases it was unknown or unrecorded; in these cases we assumed the
mosquitoes were newly emerged at the time of release and return to this
assumption later. See Table S\ref{S-tab:mrr_aggregateData} for a summary of other data characteristics.

We analysed all MRR experiments within the same statistical framework
(for full details see the Supplementary Online Material (SOM)). In the
simplest case \emph{N\textsubscript{R}} mosquitoes are released on day
zero and the probability that they remain in the recapture area until
day \emph{t} is \emph{S}(\emph{t}) when they are recaptured with
probability $\psi$. We model the number of mosquitoes recaptured on
day \emph{t} using a negative binomial sampling model with mean
\(\left( N_{R} - Y\left( t - 1 \right) \right)S\left( t \right)\psi\),
where \(Y\left( t - 1 \right)\) is cumulative captures before day
\emph{t}, and shape parameter $\kappa$. The negative binomial has been
used previously in analyses of mosquito count data \citep{service1971studies,nedelman1983negative} because of its ability to
represent temporal over-dispersion in recaptures most likely caused by
variable weather. A slight modification was required for studies with
multiple releases (see SOM).

The simplest model for $S(t)$ assumes there is a constant
probability ($\lambda$) that a mosquito dies or leaves the recapture area
so that the numbers remaining after time \emph{t} are given by the
exponential distribution, $\text{exp}( -\lambda t )$.
We utilised this form extensively but in testing for senescence used
five other models where $\lambda(t)$ varies with time so that,

\begin{longtable}[]{@{}ll@{}}
\(S\left( t \right) = e^{- \int_{0}^{t}{\lambda\left( \tau \right) \mathrm{d}\tau}}.\)
\end{longtable}

@@

Parameters were estimated using Bayesian techniques with relative
uninformative priors for $\kappa$ and the parameters of
$\lambda(t)$, but assuming a prior for $\psi$ indicating a low
recapture probability (bounded in part by knowledge of the maximum daily
recapture rates; see SOM). We used a Bayesian hierarchical model to estimate
distributions of lifespan at the species and the genus levels, and
across the complete data set. This procedure assumes that there is a
distribution of lifespan parameters for each species from which those
governing individual MRR time series are sampled, and similarly a
distribution at the genus level from which those for individual species
are derived (rather akin to random effects in classical statistics).
Within this framework we can also allow the parameters for individual
time series to be influenced by co-variates such as differences in
experimental methodology. As in the estimation of the parameters of the
individual experiments, relative uninformative priors were set for the
parameters of the hierarchical models except for $\psi$ where again a
distribution representing low recapture probabilities was assumed.
Posterior distributions were derived using Markov Chain Monte Carlo
(MCMC) methods with convergence assessed using the \(\hat{R}\) statistic
\citep{gelman1992inference}. The predictive power of the model was assessed
using \emph{K}-fold cross validation which tests the ability of the
model fitted to part of the data to predict the rest using multiple
different partitions. Further details of the prior specification,
fitting and validation through posterior predictive checks are given in
the SOM.

@@

The Guerra et al., (2014) database included the latitude and longitude of each
study along with the date when the study began. We used this information to find
estimates of the air temperature for each study using the European Centre for Medium
Range Weather Forecasts' ERA Interim Daily historical database. For each
study we calculated the mean monthly temperature across a spatial area
of (latitude $\pm$ 1 degree, longitude $\pm$ 1 degree), for the month at which
each study was carried out. The records for this database begin in 1979,
which pre-dates the study date for 65 of our 232 MRR time-series. For
these time-series, we chose to estimate the air temperature by an
average of the corresponding monthly temperatures over the years
1979-89.

@@

Most dissection studies recorded the distribution of the number of
gonotrophic cycles in mosquito samples collected over a specific period
of time. Overall, we found 568 time series in 72 published articles.
Because seldom were sufficient ancillary data available to analyse
trends over time, we aggregated the data from each time series. We
further omitted time series with fewer than 100 mosquitoes and for
species with only one data set leaving 131 studies of mosquitoes in the
genera \emph{Anopheles, Aedes}, \emph{Culex} and \emph{Mansonia}.

@@

We modelled the number of mosquitoes found by dissection to be of age
\emph{a} using the negative binomial distribution with mean
$\Psi S(a)$ and shape parameter $\kappa$, where
$\Psi$ is the product of the recruitment rate of adult mosquitoes,
which we assume is constant over time, and the probability of being
captured for dissection, and \emph{S}(\emph{a}) is the probability of
surviving until age \emph{a}. We used the number of females that have
yet to lay eggs (nulliparous) to estimate the recruitment rate as
described further in the SOM. Initial examination revealed that in some
data sets the number of nulliparous females was anomalously low,
something that has been noticed before \citep{gillies1965study}. As some
studies have suggested that the first gonotrophic cycle tends to be
longer than the subsequent ones, this is probably due to differences in
capture probability. In data sets where the fraction of nulliparous
females was less than 90\% the uniparous (completed on gonotrophic
cycle) we excluded the nulliparous observation.

@@

MRR estimates the length of time a mosquito remains alive and is still
in the area available for recapture. It thus provides a lower bound to mean
lifespan which we shall refer to as LBL. In 187 of the 230 MRR time
series the estimated LBL was less than 10 days (Fig. \ref{fig:mrr_lifespan_individualEstimates}). The smallest
estimate was 1.1 days for the Asian malaria vector
\emph{Anopheles subpictus s.l.} which is unfeasibly short and almost
certainly reflects dispersal out of the recapture zone or a violation of
the assumptions of our analyses. The longest estimate was 26.9 days for
the temperate species \emph{Aedes simpsoni s.l.} which is a vector of
yellow fever in Africa. There are multiple data sets for the most
important vector species such as \emph{Anopheles gambiae, Aedes aegypti}
and \emph{albopictus} and \emph{Culex tarsalis} all of which show
considerable variation. For example, there are 54 estimates of LBL for
\emph{Ae. aegypti} which range from 2.2 days to 38.3 days with a mean of
8.3 days and coefficient of variation of 0.7. There were significant
differences in LBL amongst species (ANOVA on median LBL controlling for
sex and pre-release feeding: \emph{F}\textsubscript{37,194} = 2.5,
\emph{p} \textless{}0.01; the non-parametric Kruskal Wallace:
\({\chi^{2}}_{38}\), \emph{p}\textless{}0.01).

The estimated mean LBL for female mosquitoes not pre-fed with blood or
sugar before release for \emph{Culex, Anopheles} and \emph{Aedes} were
2.9, 6.8 and 8.1 days respectively with an overall estimate of 6.0 days
(Fig. \ref{fig:mrr_lifespans}; Table S\ref{S-tab:mrr_estimated_lifespans}). Differences between genera were significant (ANOVA on median
LBL controlling for sex and pre-release feeding:
\emph{F}\textsubscript{2,229} = 12.4, \emph{p}\textless{}0.01; Kruskal
Wallace: \({\chi^{2}}_{2} = 30.8,p < 0.01\)). \emph{K}-fold cross
validation suggests that after the effect of genus is accounted for the
incorporation of a species term provides little predictive power (Fig.
S\#; in part explained by the latter model over-fitting the data where
there are few time series per species).

@@

To access whether temperature is associated with LBL we used weather
records to calculate average temperatures at the MRR sites (see
Methods). Using both linear and quadratic temperature terms in
regressions, we found no significant relationship between study-site
temperature and LBL (overall or within genus) for the 238 datasets we
analysed (Fig. \ref{fig:mrr_temperature}). This result held if, instead of pooling results from
all time series, we considered the four species with the most data
individually (\emph{Ae. aegypti}, \emph{Cx. tarsalis}, \emph{A. gambiae s.l.} and
\emph{A. culicifacies s.l.}; Fig. \ref{S-fig:mrr_ThreeSpeciesVersusTemperature}).
